---
title: "Mutational Characterization"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::html_document2:
    css: styles.css
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

### Libraries and Paths

```{r libraries-paths}
library(grid)
library(glue)
library(dplyr)
library(musicatk)
library(maftools)
library(circlize)
library(gridExtra)
library(ComplexHeatmap)
library(ggpubr)
library(rstatix)
library(ggplot2)
library(tidyr)
library(purrr)
library(readxl)
library(tibble)
library(data.table)
library(writexl)
library(ggpubr)
source("/restricted/projectnb/pcga/DNA_Seq_Combined_Analysis/Project/Scripts/readData.R")

data_dir <- "/restricted/projectnb/pcga/DNA_Seq_Combined_Analysis/Project/Data"
```

### Helper Functions

```{r helpers}
getMutTable <- function(maf, annotation, genes = driver_genes) {
  mat = mutCountMatrix(maf, removeNonMutated = FALSE)
  mat <- mat[rownames(mat) %in% genes, , drop = FALSE]
  mat = mat %>%
    t() %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var = "Tumor_Sample_Barcode")

  for (gene in genes) {
    if (!gene %in% colnames(mat)) {
      mat[[gene]] <- 0
    }
  }

  merge(
    x = mat,
    y = annotation[, c("Tumor_Sample_Barcode", "Histology")],
    by = "Tumor_Sample_Barcode",
    all.x = TRUE
  )
}

get_histology_sample_ids <- function(annotation, histologies) {
  setNames(
    lapply(histologies, function(h) annotation$Tumor_Sample_Barcode[annotation$Histology == h]),
    histologies
  )
}

subset_maf_by_histology <- function(maf, sample_ids, genes = driver_genes) {
  lapply(sample_ids, function(ids) {
    subsetMaf(maf, genes = genes, tsb = ids)
  })
}

get_sample_summary_df <- function(maf_list, cohort_sizes) {
  dat <- purrr::imap_dfr(maf_list, function(maf, hist_name) {
    sample_vars = getSampleSummary(maf)
    with_drivers = nrow(sample_vars)
    cohort_size = cohort_sizes[[hist_name]]
    tibble(
      Histology = hist_name,
      percentage = round(with_drivers / cohort_size * 100),
      with_drivers = with_drivers,
      zero_drivers = cohort_size - with_drivers
    )
  })
  dat
}

samples_with_genes <- function(maf, hist, samples) {
  mat = data.frame(t(mutCountMatrix(maf)))
  mat = mat |> mutate(across(everything(), ~ifelse(. > 0, 1, 0)))

  x = mat |> 
    rowSums() |>
    data.frame() |>
    rename(count = 1) |>
    tibble::rownames_to_column(var = "Tumor_Sample_Barcode")

  x = bind_rows(
    x,
    data.frame(
      Tumor_Sample_Barcode = samples[!samples %in% x$Tumor_Sample_Barcode],
      count = 0
    )
  )

  x$Histology = hist
  x
}

get_gene_mutation_table <- function(mut.dat, group_vec, gene_name, group_label) {
  gene_hits = substring(subset(mut.dat, Hugo_Symbol == gene_name)$Tumor_Sample_Barcode, 1, 30)
  gene_ix = names(group_vec) %in% gene_hits
  table(gene_ix, group_vec == group_label)
}

get_gene_outcome_table <- function(mut.dat, outcome_vec, gene_name) {
  gene_hits = substring(subset(mut.dat, Hugo_Symbol == gene_name)$Tumor_Sample_Barcode, 1, 30)
  gene_ix = names(outcome_vec) %in% gene_hits
  table(gene_ix, outcome_vec)
}

make_gene_barplot <- function(plot_df, stat_df, plot_title, y_headroom, p_headroom) {
  ggplot(data = plot_df, aes(x = type, y = Frequency, fill = type)) +
    geom_bar(stat = "identity", width = 0.7, alpha = 0.8) + 
    scale_fill_manual(values = c("blue4", "red3")) +
    guides(fill = "none") +
    theme_classic() +
    ylab("Fraction mutated") +
    xlab("") + 
    theme(
      axis.text.x = element_text(size = 6, angle = 45, hjust = 1, family = "Helvetica", color = "black"), 
      axis.text.y = element_text(size = 5, family = "Helvetica", color = "black"), 
      axis.title.y = element_text(size = 6.5, vjust = 2.5, family = "Helvetica"), 
      axis.ticks.x = element_line(linewidth = 0.25), 
      axis.ticks.y = element_line(linewidth = 0.25),
      plot.title = element_text(size = 6.5, family = "Helvetica", hjust = 0.5, vjust = 1, face = "italic"), 
      plot.margin = unit(c(0, 0, -0.5, 0.25), "cm"),
      axis.line.x = element_line(linewidth = 0.3),
      axis.line.y = element_line(linewidth = 0.3)
    ) +
    ggtitle(plot_title) + 
    scale_y_continuous(
      limits = c(0, max(plot_df$Frequency) + y_headroom),
      expand = c(0, 0)
    ) +
    stat_pvalue_manual(
      data = stat_df,
      label = "p = {p_val}",
      y.position = max(plot_df$Frequency) + p_headroom,
      size = 2,
      inherit.aes = FALSE
    )
}

plotTMBplot <- function(plot.dat, tmb.merge, wilcox.res, histology, logscale = TRUE) {
  if (logscale) {
    y_lims = c(-2, 2.3)
    y_at = log10(c(0.01, 0.1, 1, 10, 100))
  } else {
    y_lims = range(data.table::rbindlist(l = plot.dat)[, total_perMB])
    y_at = pretty(y_lims)
  }

  tmb.merge = data.frame(tmb.merge)
  col = c("grey0", "black")
  bg_col = c("white", "lightgrey")
  medianCol = "red3"
  cohortFontSize = 0.5
  axisFontSize = 0.5

  plot(NA, NA, xlim = c(0, length(histology)), ylim = y_lims, axes = FALSE, xlab = NA, ylab = NA, frame.plot = TRUE)
  rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], col = grDevices::adjustcolor(col = "gray", alpha.f = 0.1))
  rect(
    xleft = seq(0, length(histology) - 1, 1),
    ybottom = min(y_lims),
    xright = seq(1, length(histology), 1),
    ytop = max(y_lims) + 5,
    col = grDevices::adjustcolor(col = bg_col, alpha.f = 0.2),
    border = NA
  )
  abline(h = pretty(y_lims), lty = 2, col = "gray70")

  lapply(seq_len(length(plot.dat)), function(i) {
    x = plot.dat[[i]]
    if (logscale) {
      points(x$V1, log10(x$total_perMB), pch = 16, cex = 0.25, col = col[1])
    } else {
      points(x$V1, x$total_perMB, pch = 16, cex = 0.25, col = col[1])
    }
  })

  axis(side = 1, at = seq(0.5, length(plot.dat) - 0.5, 1), labels = FALSE, tick = FALSE)
  text(
    x = seq(0.5, length(plot.dat) - 0.5, 1),
    y = par("usr")[3] - 0.20,
    labels = names(plot.dat),
    srt = 45,
    adj = 1,
    xpd = TRUE,
    cex = cohortFontSize
  )

  sidePos = 2
  linePos = 2

  if (logscale) {
    axis(side = 2, at = y_at, las = 2, line = -0.6, tick = FALSE, labels = round(10^(y_at), 5), cex.axis = axisFontSize)
  } else {
    axis(side = 2, at = y_at, las = 2, line = -0.6, tick = FALSE, cex.axis = axisFontSize)
  }
  mtext(text = "Mutations per Megabase", side = sidePos, line = linePos, cex = 0.6)

  if (logscale) {
    tmb.merge$Median_Mutations_log10 <- log10(tmb.merge$Median_Mutations)
    lapply(seq_len(nrow(tmb.merge)), function(i) {
      segments(
        x0 = i - 1, x1 = i,
        y0 = tmb.merge[i, "Median_Mutations_log10"],
        y1 = tmb.merge[i, "Median_Mutations_log10"],
        col = medianCol, lwd = 1
      )
    })

    text(
      x = wilcox.res$group2 - 0.5,
      y = wilcox.res$ypos,
      labels = wilcox.res$p.adj.signif,
      cex = 0.4,
      xpd = TRUE
    )

    segments(
      x0 = 1.5,
      x1 = wilcox.res$group2 - 0.5,
      y0 = wilcox.res$ypos - 0.1,
      y1 = wilcox.res$ypos - 0.1,
      lwd = 0.75
    )

    segments(
      x0 = 1.5, x1 = 1.5,
      y0 = wilcox.res$ypos - 0.15, y1 = wilcox.res$ypos - 0.1,
      lwd = 0.75
    )

    segments(
      x0 = wilcox.res$group2 - 0.5, x1 = wilcox.res$group2 - 0.5,
      y0 = wilcox.res$ypos - 0.15, y1 = wilcox.res$ypos - 0.1,
      lwd = 0.75
    )
  } else {
    lapply(seq_len(nrow(tmb.merge)), function(i) {
      segments(
        x0 = i - 1, x1 = i,
        y0 = tmb.merge[i, "Median_Mutations"],
        y1 = tmb.merge[i, "Median_Mutations"],
        col = medianCol
      )
    })
  }
}
```

### Load Input Data and Shared Objects

```{r load-data}
pcga.file <- read.delim(file.path(data_dir, "20200201_PCGA_Exome_with_MuTect2PON.maf"))
pcga.file$Tumor_Sample_Barcode <- substr(pcga.file$Tumor_Sample_Barcode, 1, 30)
pcga.annotations <- read_xlsx(file.path(data_dir, "20240802-PCGA-01-Metadata-New.xlsx"))
pcga.annotations <- pcga.annotations[pcga.annotations$Tumor_Sample_Barcode %in% pcga.file$Tumor_Sample_Barcode,]

cis.annotations <- read_xlsx(file.path(data_dir, "pheno.for.josh.xlsx"))
cis.annotations <- cis.annotations |>
  rename(Tumor_Sample_Barcode = Sample.Number..WGS.) |>
  mutate(Histology = "CIS")

cis.file <- read.delim(file.path(data_dir, "CIS_maftools.maf"), quote = "")

tcga.data <- read.delim(
  file.path(data_dir, "tcga.hg19.maf"),
  sep = "\t",
  quote = "",
  fill = TRUE,
  stringsAsFactors = FALSE
)

tcga.annotations <- data.frame(
  Tumor_Sample_Barcode = unique(tcga.data$Tumor_Sample_Barcode),
  Histology = "LUSC"
)

driver_genes <- read.table(file.path(data_dir, "lusc_genes.txt"))[, 1]

pcga_histologies <- c("Brush", "Normal", "Hyperplasia", "Metaplasia", "Mild Dysplasia", "Moderate Dysplasia", "Severe Dysplasia")
histology_no_brush <- c("Normal", "Hyperplasia", "Metaplasia", "Mild Dysplasia", "Moderate Dysplasia", "Severe Dysplasia", "CIS", "LUSC")
histology_all <- c("Brush", histology_no_brush)
outcome_levels <- c("stable", "regressive", "progressive", "persistent")

histCols <- c("skyblue", "grey45", "#3c68b0", "#F4AD31", "#FC9272", "#FB6A4A", "#DE2D26")
names(histCols) <- pcga_histologies

histCols_no_brush <- c("grey45", "#3c68b0", "#F4AD31", "#FC9272", "#FB6A4A", "#DE2D26", "#7B2D3E", "#2E1410")
names(histCols_no_brush) <- histology_no_brush

histCols_all <- c("skyblue", histCols_no_brush)
names(histCols_all) <- histology_all

outcomeCols <- c("#E0E0E0", "#B0BEC5", "#00796B", "#FB8C00", "#cd7058")
names(outcomeCols) <- c("unknown", "stable", "regressive", "progressive", "persistent")

smoke.cols <- c("#4169E1", "#CD5C5C", "#F8F8FF")
names(smoke.cols) <- c("Former", "Current", "NA")

driver_variants <- c("Missense_Mutation", "Nonsense_Mutation", "Nonstop_Mutation", "Splice_Site")
signature_variants <- c("Missense_Mutation", "Nonsense_Mutation", "Nonstop_Mutation", "Silent", "Splice_Site")

pcga.maf_full <- ReadMaf(maf = pcga.file, metadata = pcga.annotations)

pcga.maf_driver <- read.maf(pcga.file, clinicalData = pcga.annotations, vc_nonSyn = driver_variants)
cis.maf_driver <- read.maf(cis.file, clinicalData = cis.annotations, vc_nonSyn = driver_variants)
tcga.maf_driver <- read.maf(tcga.data, vc_nonSyn = driver_variants, clinicalData = tcga.annotations)

pcga_sample_ids <- get_histology_sample_ids(pcga.annotations, pcga_histologies[pcga_histologies != 'Brush'])
pcga_driver_mafs_by_histology <- subset_maf_by_histology(pcga.maf_driver, pcga_sample_ids, genes = driver_genes)
names(pcga_driver_mafs_by_histology) <- pcga_histologies[pcga_histologies != 'Brush']

cis.maf_driver_subset <- subsetMaf(cis.maf_driver, genes = driver_genes)
tcga.maf_driver_subset <- subsetMaf(tcga.maf_driver, genes = driver_genes)

cohort_sizes <- c(
  Normal = 39,
  Hyperplasia = 49,
  Metaplasia = 21,
  `Mild Dysplasia` = 17,
  `Moderate Dysplasia` = 26,
  `Severe Dysplasia` = 12,
  CIS = 39,
  LUSC = 473
)

update_geom_defaults("text", list(family = "Helvetica"))
```

### Somatic Mutational Landscape of PCGA

```{r oncoplot, fig.width=8, fig.height=7, fig.dpi=300, fig.align='center', fig.cap="Figure 1B", out.width='100%', message=FALSE, warning=FALSE}
nsamples <- length(unique(pcga.file$Tumor_Sample_Barcode))
tmb_dat <- tmb(pcga.maf_full, captureSize = 30, plotType = NA) |> data.frame() |> select(c("Tumor_Sample_Barcode", "total_perMB"))
colnames(tmb_dat) <- c("Tumor_Sample_Barcode", "Mutations per Mb")

mut_count_mat <- maftools::mutCountMatrix(pcga.maf_full, removeNonMutated = FALSE)
mut_count_mat <- ifelse(mut_count_mat > 0, 1, 0)

gen_percent <- round(rowSums(mut_count_mat) / nsamples * 100)
gen_percent <- gen_percent[gen_percent > 0]
most_mut_genes <- names(gen_percent)[names(gen_percent) %in% driver_genes]

annCols <- list(
  Histology = histCols,
  Outcome = outcomeCols,
  Genomic_Smoking_Status = smoke.cols
)

vc_cols <- c(
  RColorBrewer::brewer.pal(n = 12, name = "Paired"),
  RColorBrewer::brewer.pal(n = 12, name = "Set3")[1]
)
names(vc_cols) <- unique(pcga.maf_full@data$Variant_Classification)

oncoplot(
  pcga.maf_full,
  genes = most_mut_genes,
  clinicalFeatures = c("Histology", "Outcome", "Genomic_Smoking_Status"),
  sortByAnnotation = FALSE,
  bgCol = "white",
  showTitle = TRUE,
  removeNonMutated = TRUE,
  borderCol = "#EEEEEE",
  annotationColor = annCols,
  drawBox = TRUE,
  gene_mar = 7,
  topBarData = tmb_dat,
  sepwd_genes = 1.5,
  sepwd_samples = 1.5,
  colors = vc_cols,
  topBarLims = c(0, 8),
  drawRowBar = FALSE,
  anno_height = 0.75,
  annotationFontSize = 0.6, 
  legendFontSize = 0.6,
  titleFontSize = 1, 
  fontSize = 0.6
)
```

### Percent of Samples With Driver Mutations

```{r samples-with-drivers, fig.width=2, fig.height=2.5, fig.align='center', fig.dpi=300, fig.cap="Figure 1C", message=FALSE, warning=FALSE}
driver_maf_list <- c(
  pcga_driver_mafs_by_histology[histology_no_brush[histology_no_brush %in% names(pcga_driver_mafs_by_histology)]],
  list(CIS = cis.maf_driver_subset, LUSC = tcga.maf_driver_subset)
)

driver_summary <- get_sample_summary_df(driver_maf_list, cohort_sizes)
driver_summary$Histology <- factor(driver_summary$Histology, levels = histology_no_brush)

mut <- as.matrix(driver_summary[, c("with_drivers", "zero_drivers")])
rownames(mut) <- driver_summary$Histology
ft <- fisher.test(mut, simulate.p.value = TRUE)
p_label <- paste0("Fisher's exact test\np = ", signif(ft$p.value, 3))

ggplot(driver_summary, aes(x = Histology, y = percentage, fill = Histology)) +
  geom_col() +
  coord_cartesian(ylim = c(0, 100)) +
  annotate("text", x = Inf, y = Inf, label = p_label, size = 1, hjust = 1.1, vjust = 1.3) +
  theme_bw() +
  xlab("") +
  ylab("% of samples with driver mutation") +
  scale_fill_manual(values = histCols_no_brush) +
  theme(
    axis.text.x = element_text(size = 5, family = "Helvetica", color = "black", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5, family = "Helvetica", color = "black"),
    axis.title.y = element_text(size = 6, family = "Helvetica"),
    legend.position = "none",
    panel.grid = element_blank()
  )
```

### Number of Driver Genes Mutated Per Sample

```{r drivers-per-sample, fig.align='center', fig.width=2, fig.height=2.5, fig.dpi=300, fig.cap="Figure 1D", message=FALSE, warning=FALSE}
norm.samp <- pcga_sample_ids[["Normal"]]
hyper.samp <- pcga_sample_ids[["Hyperplasia"]]
meta.samp <- pcga_sample_ids[["Metaplasia"]]
mild.samp <- pcga_sample_ids[["Mild Dysplasia"]]
mod.samp <- pcga_sample_ids[["Moderate Dysplasia"]]
sev.samp <- pcga_sample_ids[["Severe Dysplasia"]]
cis.samp <- unique(cis.file$Tumor_Sample_Barcode)
tcga.samp <- unique(tcga.data$Tumor_Sample_Barcode)

df_driver_counts <- bind_rows(
  samples_with_genes(pcga_driver_mafs_by_histology[["Normal"]], "Normal", norm.samp),
  samples_with_genes(pcga_driver_mafs_by_histology[["Hyperplasia"]], "Hyperplasia", hyper.samp),
  samples_with_genes(pcga_driver_mafs_by_histology[["Metaplasia"]], "Metaplasia", meta.samp),
  samples_with_genes(pcga_driver_mafs_by_histology[["Mild Dysplasia"]], "Mild Dysplasia", mild.samp),
  samples_with_genes(pcga_driver_mafs_by_histology[["Moderate Dysplasia"]], "Moderate Dysplasia", mod.samp),
  samples_with_genes(pcga_driver_mafs_by_histology[["Severe Dysplasia"]], "Severe Dysplasia", sev.samp),
  samples_with_genes(cis.maf_driver_subset, "CIS", cis.samp),
  samples_with_genes(tcga.maf_driver_subset, "LUSC", tcga.samp)
)

df_driver_counts$Histology <- factor(df_driver_counts$Histology, levels = histology_no_brush)
kruskal_driver_counts <- kruskal.test(count ~ Histology, data = df_driver_counts)

wilcox.res <- pairwise.wilcox.test(
  x = df_driver_counts$count,
  g = df_driver_counts$Histology,
  p.adjust.method = "bonferroni",
  exact = FALSE
)[[3]]

wilcox.res <- wilcox.res |>
  data.frame() |>
  rownames_to_column(var = "group2") |>
  pivot_longer(cols = !group2, names_to = "group1", values_to = "p") |>
  relocate(group1) |>
  filter(!is.na(p)) |>
  mutate(
    p.signif = case_when(
      p < 0.0001 ~ "****",
      p < 0.001 ~ "***",
      p < 0.01 ~ "**",
      p < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

wilcox.res$group1 <- gsub("\\.", " ", wilcox.res$group1)
wilcox.res <- wilcox.res %>%
  add_xy_position(data = df_driver_counts, formula = count ~ Histology, x = "Histology")

wilcox.res <- wilcox.res[, -5]
y_max <- df_driver_counts |>
  group_by(Histology) |>
  summarise(y.position = max(count) + 1) |>
  rename(group2 = Histology)

wilcox.res <- merge(wilcox.res, y_max, by = "group2")

ggplot(df_driver_counts, aes(x = Histology, y = count, fill = Histology)) +
  geom_boxplot(outlier.size = 0.01, outlier.colour = "black", size = 0.3, lwd = 0.2) +
  geom_jitter(size = 0.01, height = 0.02) +
  theme_bw() +
  xlab("") +
  ylab("# driver genes mutated per sample") +
  scale_y_continuous(breaks = c(0, 2, 4, 6, 8)) +
  scale_fill_manual(values = histCols_no_brush) +
  theme(
    axis.text.x = element_text(size = 5, family = "Helvetica", color = "black", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5, family = "Helvetica", color = "black"),
    axis.title.y = element_text(size = 5, family = "Helvetica"),
    legend.position = "none",
    panel.grid = element_blank()
  ) +
  stat_pvalue_manual(
    data = wilcox.res,
    tip.length = 0.01,
    xmin = "group1",
    xmax = "group2",
    label = "p.signif",
    label.size = 1.5,
    inherit.aes = FALSE,
    hide.ns = TRUE
  ) +
  annotate(
    geom = "text",
    label = paste0("Kruskal-Wallis test p-value = ", format(kruskal_driver_counts$p.value, scientific = TRUE, digits = 2)), size = 1,
    y = 8.5,
    x = 4
  )
```

### Mutation Frequencies of Driver Genes Across PCGA, CIS, and TCGA

```{r mutation-freq, fig.width=4, fig.height=4,fig.align='center', fig.dpi=300, fig.cap="Figure 1E", message=FALSE, warning=FALSE}
mut <- bind_rows(
  getMutTable(pcga.maf_driver, pcga.annotations),
  getMutTable(cis.maf_driver, cis.annotations),
  getMutTable(tcga.maf_driver, tcga.annotations)
)

mut$Histology <- factor(mut$Histology, levels = histology_all)
mut <- mut[!(is.na(mut$Histology)), ]

set.seed(123)
results <- list()
for (i in seq_along(driver_genes)) {
  mat = t(table(mut[driver_genes[i]] > 0, mut$Histology))
  results[[driver_genes[i]]] <- fisher.test(mat, alternative = "greater", simulate.p.value = TRUE)$p.value
}

res.df <- tibble::enframe(results, name = "Genes", value = "p.value")
res.df$p.value <- unlist(res.df$p.value)
res.df$fdr <- p.adjust(res.df$p.value, method = "fdr")

m <- mut %>%
  pivot_longer(
    cols = colnames(mut)[colnames(mut) %in% driver_genes],
    names_to = "Genes",
    values_to = "Mutation"
  )

m_long <- m %>%
  group_by(Histology, Genes) %>%
  summarize(Mutated = sum(Mutation != 0, na.rm = TRUE), .groups = "drop") %>%
  filter(complete.cases(.))

matx <- m_long %>% pivot_wider(names_from = "Histology", values_from = "Mutated")
matx$highly_mut_genes <- rowSums(matx[, -1])
matx <- matx[order(matx$highly_mut_genes, decreasing = TRUE), ]
matx <- matx[, -11]

gene.row <- matx$Genes
matx <- matx[, histology_all]

total_sizes <- data.frame(table(mut$Histology))[, 2]

plot_data <- as.data.frame(mapply(function(x, y) x / y, matx, total_sizes)) * 100
rownames(plot_data) <- gene.row
plot_data <- plot_data[, -1]

col_fun <- colorRamp2(c(0, 15, 75), c("white", "#F05039", "#1F449C"))
plt <- Heatmap(
  as.matrix(plot_data),
  name = " ",
  col = col_fun,
  rect_gp = gpar(col = "lightgrey", lwd = 0.25),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  heatmap_legend_param = list(
    at = c(0, 15, 30, 75),
    legend_height = unit(1.5, "cm"),
    grid_width = unit(0.25, "cm"),
    labels_gp = gpar(fontsize = 5)
  ),
  column_order = names(plot_data),
  row_order = rownames(plot_data),
  row_names_gp = gpar(fontsize = 6, fontfamily = "Helvetica", fontface = "italic"),
  row_names_side = "left",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 6, fontfamily = "Helvetica")
)

heatmap <- grid.grabExpr(draw(plt, heatmap_legend_side = "left"))

res.df <- res.df %>%
  mutate(Genes = factor(Genes, levels = gene.row)) %>%
  arrange(Genes) %>%
  as.data.frame()

rownames(res.df) <- c(25:1)
cutoff <- -log10(0.05)

bp <- res.df %>%
  ggplot(aes(x = forcats::fct_reorder(Genes, as.numeric(rownames(res.df))), y = -log10(fdr))) +
  geom_col(fill = "grey45", width = 0.75) +
  coord_flip() +
  theme_classic() +
  geom_errorbar(aes(ymin = 0, ymax = 0), width = 0.75, color = "grey45", lwd = 0.5) +
  xlab("") +
  ylab("-log10(q-value)") +
  geom_hline(yintercept = cutoff, linetype = "dashed", color = "red", lwd = 0.5) +
  theme(
    axis.text.x = element_text(family = "Helvetica", size = 6, colour = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 6),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank()
  )

grid.arrange(
  heatmap,
  bp,
  ncol = 2,
  widths = unit(c(1, 1), "null"),
  heights = unit(c(1), "null"),
  padding = unit(0.1, "cm")
)
```

### NOTCH1 in Dysplasia vs Low-Grade Lesions

```{r dys-vs-lowgrade, fig.width=2.5, fig.height=2, fig.align='center', fig.dpi=300, fig.cap="Figure 1F", message=FALSE, warning=FALSE}
mut.dat <- pcga.maf_full@data
pcga.clinc <- pcga.maf_full@clinical.data

histology_grouped <- setNames(pcga.clinc$Histology, pcga.clinc$Tumor_Sample_Barcode)
histology_grouped[histology_grouped %in% c("Severe Dysplasia", "Moderate Dysplasia", "Mild Dysplasia")] <- "Dysplasia"

notch1_dys <- get_gene_mutation_table(mut.dat, histology_grouped, "NOTCH1", "Dysplasia")
tp53_dys <- get_gene_mutation_table(mut.dat, histology_grouped, "TP53", "Dysplasia")

stat1 <- data.frame(group1 = "Other", group2 = "Dysplasia", p_val = round(fisher.test(notch1_dys, alternative = "greater")$p.value, 3))
stat2 <- data.frame(group1 = "Other", group2 = "Dysplasia", p_val = round(fisher.test(tp53_dys, alternative = "greater")$p.value, 3))

plt1 <- data.frame(type = c("Other", "Dysplasia"), Frequency = c(t(prop.table(notch1_dys, 2)[2, ])[[1]], t(prop.table(notch1_dys, 2)[2, ])[[2]]))
plt1$type <- factor(plt1$type, levels = c("Other", "Dysplasia"))

plt2 <- data.frame(type = c("Other", "Dysplasia"), Frequency = c(t(prop.table(tp53_dys, 2)[2, ])[[1]], t(prop.table(tp53_dys, 2)[2, ])[[2]]))
plt2$type <- factor(plt2$type, levels = c("Other", "Dysplasia"))

notch1.barplot <- make_gene_barplot(plt1, stat1, "NOTCH1", 0.04, 0.02)
tp53.barplot <- make_gene_barplot(plt2, stat2, "TP53", 0.02, 0.01)

ggarrange(notch1.barplot, tp53.barplot, ncol = 2, nrow = 1)
```

### NOTCH1 Mutations by Lesion Outcome

```{r mutation-by-outcome, fig.width=3, fig.height=2, fig.align='center', fig.dpi=300, fig.cap="Figure 1G", message=FALSE, warning=FALSE}
pcga.clinc_outcome <- pcga.maf_full@clinical.data
pcga.clinc_outcome <- pcga.clinc_outcome[pcga.clinc_outcome$Outcome != "unknown", ]
outcome <- setNames(pcga.clinc_outcome$Outcome, pcga.clinc_outcome$Tumor_Sample_Barcode)

notch1_outcome <- get_gene_outcome_table(mut.dat, outcome, "NOTCH1")
tp53_outcome <- get_gene_outcome_table(mut.dat, outcome, "TP53")

plt1 <- data.frame(t(prop.table(notch1_outcome, 2)[2, ])) |>
  pivot_longer(cols = everything(), names_to = "outcome", values_to = "Frequency") |>
  mutate(gene = "NOTCH1")

plt2 <- data.frame(t(prop.table(tp53_outcome, 2)[2, ])) |>
  pivot_longer(cols = everything(), names_to = "outcome", values_to = "Frequency") |>
  mutate(gene = "TP53")

df <- bind_rows(plt1, plt2)
df$outcome <- factor(df$outcome, levels = outcome_levels)

stat <- data.frame(
  gene = c("NOTCH1", "TP53"),
  pVal = c(
    fisher_test(notch1_outcome, alternative = "greater")$p,
    fisher_test(tp53_outcome, alternative = "greater")$p
  )
)
ggplot(df, aes(x = outcome, y = Frequency, fill = outcome)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(
    data = stat,
    mapping = aes(x = 0.5, y = 0.45, label = paste0("p=", pVal)),
    size = 2,
    hjust = -0.1,
    vjust = 0,
    inherit.aes = FALSE
  ) +
  facet_grid(~gene, scales = "free_y") +
  theme_classic() +
  scale_fill_manual(values = c("#B0BEC5", "#00796B", "#FB8C00", "#cd7058")) +
  ylab("Fraction mutated") +
  xlab("") +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_line(linewidth = 0.25),
    axis.text.y = element_text(size = 6, family = "Helvetica", color = "black"),
    axis.title.y = element_text(size = 6, vjust = 2.5, family = "Helvetica"),
    plot.title = element_text(size = 6, family = "Helvetica"),
    strip.text = element_text(size = 6, family = "Helvetica", face = "italic"),
    plot.margin = unit(c(0, 0, 0, 0.25), "cm"),
    strip.background = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size = 4),
    legend.key.size = unit(0.2, "cm"),
    legend.position = "bottom",
    legend.box.spacing = unit(0, "pt"),
    legend.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(0.1, "cm"),
    axis.line.x = element_line(linewidth = 0.3),
    axis.line.y = element_line(linewidth = 0.3)
  ) +
  guides(fill = guide_legend(nrow = 1)) +
  geom_errorbar(aes(ymin = 0, ymax = 0, color = outcome), width = 0.7, show.legend = FALSE) +
  scale_color_manual(values = c("#B0BEC5", "#00796B", "#FB8C00", "#cd7058"))
```

### Tumor Mutation Burden

```{r TMB, fig.width=4, fig.height=4.5, fig.dpi=300, fig.align='center', fig.cap="Figure 1H", message=FALSE, warning=FALSE}
tmb_list <- list()
exome.clinic <- getClinicalData(pcga.maf_driver)

for (hist in pcga_histologies) {
  sample_ids <- exome.clinic[Histology %in% hist, Tumor_Sample_Barcode]
  sub.maf <- subsetMaf(pcga.maf_driver, tsb = sample_ids, verbose = FALSE)
  tmb_list[[hist]] <- tmb(sub.maf, captureSize = 30, plotType = NA, verbose = FALSE)
}

tmb_list[["LUSC"]] <- tmb(tcga.maf_driver, captureSize = 30, plotType = NA, verbose = FALSE)
tmb_list[["CIS"]] <- tmb(cis.maf_driver, captureSize = 30, plotType = NA, verbose = FALSE)

for (i in seq_along(tmb_list)) {
  tmb_list[[i]]$Histology <- names(tmb_list[i])
}

mutation_load <- do.call(rbind, tmb_list) %>% select(c(5, 3))

wilcox.res <- mutation_load |>
  wilcox_test(total_perMB ~ Histology, ref.group = "Normal", p.adjust.method = "bonferroni")

wilcox.res$ypos <- c(1.0, 2.2, 1.2, 2.4, 1.4, 1.6, 1.8, 2.0)
wilcox.res$group2 <- as.numeric(factor(wilcox.res$group2, levels = unique(histology_all)))

median <- mutation_load %>% group_by(Histology) %>% summarise(Median_Mutations = median(total_perMB))
tmb.merge <- merge(data.frame(table(mutation_load[, 1])), median, by.x = "Histology", by.y = "Histology")
tmb.merge <- tmb.merge[match(histology_all, tmb.merge$Histology), ]
colnames(tmb.merge) <- c("Histology", "Cohort_Size", "Median_Mutations")

cohort.dat <- split(mutation_load, as.factor(mutation_load$Histology))
cohort.dat <- cohort.dat[histology_all]

plot.dat <- lapply(seq_len(length(cohort.dat)), function(i) {
  x = cohort.dat[[i]]
  data.table(
    rev(seq(i - 1, i, length.out = nrow(x))),
    x[order(x$total_perMB, decreasing = TRUE), "total_perMB"],
    x[, "Histology"]
  )
})

names(plot.dat) <- histology_all
plotTMBplot(plot.dat, tmb.merge, wilcox.res, histology_all, TRUE)
```

### Mutation Signature Analysis

```{r signature-analysis, fig.width=4.25, fig.height=2.5, fig.dpi=300, fig.cap="Figure 1I",fig.align='center', message=FALSE, warning=FALSE}
pcga.sub.file <- pcga.file[pcga.file$Variant_Classification %in% signature_variants, ]
cis.sub.file <- cis.file[cis.file$Variant_Classification %in% signature_variants, ]
tcga.sub.data <- tcga.data[tcga.data$Variant_Classification %in% signature_variants, ]

pcga.maf_signature <- read.maf(pcga.sub.file, clinicalData = pcga.annotations)
pcga_variants <- extract_variants_from_maf(pcga.maf_signature)
pcga_variants$chr <- paste0("chr", pcga_variants$chr)

cis.maf_signature <- read.maf(cis.sub.file, clinicalData = cis.annotations)
cis_variants <- extract_variants_from_maf(cis.maf_signature)
cis_variants$chr <- paste0("chr", cis_variants$chr)

tcga.maf_signature <- read.maf(tcga.sub.data, clinicalData = tcga.annotations)
tcga_variants <- extract_variants_from_maf(tcga.maf_signature)
tcga_variants$chr <- paste0("chr", tcga_variants$chr)

variants <- rbind(pcga_variants, cis_variants, tcga_variants)

annotations <- getClinicalData(pcga.maf_signature) %>% select(Tumor_Sample_Barcode, Histology)
comb_annot <- rbind(
  annotations,
  cis.annotations[, c("Tumor_Sample_Barcode", "Histology")],
  tcga.annotations
)
names(comb_annot) <- c("sample", "condition")

hg19 <- select_genome("hg19")
music.obj <- create_musica(variants, g = hg19, convert_dbs = FALSE)

samples <- droplevels(music.obj@sample_annotations)
samples <- as.vector(samples$Samples)
comb_annot <- comb_annot[match(samples, comb_annot$sample), ]

samp_annot(music.obj, "condition") <- comb_annot$condition
build_standard_table(musica = music.obj, g = hg19, table_name = "SBS96")

music.sub <- subset_musica_by_counts(musica = music.obj, table_name = "SBS96", num_counts = 10)

data("cosmic_v2_sigs")

result_pred_cosmic <- predict_exposure(
  musica = music.sub,
  table_name = "SBS96",
  signature_res = cosmic_v2_sigs,
  algorithm = "lda",
  signatures_to_use = c(1, 2, 4, 5, 13)
)

mut.counts <- result_pred_cosmic@exposures
mut.counts <- data.frame(t(mut.counts)) |>
  tibble::rownames_to_column(var = "Samples") |>
  pivot_longer(cols = !"Samples", names_to = "signature", values_to = "count")

mut.counts.dt <- merge(mut.counts, result_pred_cosmic@musica@sample_annotations, by = "Samples", all.x = TRUE)
wilcox.res <- mut.counts.dt |>
  group_by(signature) |>
  wilcox_test(count ~ condition, ref.group = "Normal", p.adjust.method = "bonferroni") |>
  add_xy_position(x = "condition")

exposures <- result_pred_cosmic@exposures
exposures <- sweep(exposures, 2, colSums(exposures), FUN = "/")
expos.frac <- exposures %>% t() %>% data.frame() %>% tibble::rownames_to_column(var = "Samples")
expos.frac <- expos.frac %>% pivot_longer(!Samples, names_to = "signature", values_to = "exposure")
expos.frac <- merge(expos.frac, result_pred_cosmic@musica@sample_annotations, by = "Samples", all.x = TRUE)
expos.frac$condition <- factor(expos.frac$condition, levels = histology_no_brush)

y_pos <- expos.frac %>%
  group_by(signature, condition) %>%
  summarise(ymax = max(exposure) + 0.1, .groups = "drop") %>%
  rename(group2 = condition)

wilcox.res <- left_join(wilcox.res, y_pos, by = c("signature", "group2")) |>
  select(!y.position) |>
  rename(y.position = ymax)

selected_signatures <- c("SBS4", "SBS2", "SBS13")
wilcox.res <- wilcox.res |> filter(signature %in% selected_signatures)
expos.frac <- expos.frac[expos.frac$signature %in% wilcox.res$signature, ]

expos.frac$signature <- factor(expos.frac$signature, levels = selected_signatures)
wilcox.res$signature <- factor(wilcox.res$signature, levels = selected_signatures)

ggplot(expos.frac, aes(x = condition, y = exposure, fill = condition)) +
  geom_boxplot(outlier.size = 0.05, fatten = 0.5, lwd = 0.25, show.legend = FALSE) +
  geom_point(aes(fill = condition), shape = 22, size = 0, colour = NA) +
  scale_fill_manual(values = histCols_no_brush) +
  facet_wrap(. ~ signature, nrow = 1, strip.position = "top", scales = "free_y") +
  theme_bw() +
  ylab("Exposure fractions") +
  theme(axis.text.x = element_blank()) +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 5, family = "Helvetica", color = "black"),
    axis.title.y = element_text(size = 6, family = "Helvetica"),
    legend.text = element_text(size = 4, family = "Helvetica"),
    legend.key.size = unit(0.3, "cm"),
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5, vjust = 1, size = 6, family = "Helvetica"),
    strip.background = element_blank(),
    strip.text = element_text(size = 6, family = "Helvetica", colour = "black"),
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.spacing.x = unit(1, "cm"),
    legend.key.height = unit(0.1, "cm"),
    legend.key.width = unit(0.15, "cm"),
    legend.key.spacing.y = unit(0.05, "cm"),
    panel.background = element_blank(),
    panel.border = element_rect(linewidth = 0.5),
    panel.grid = element_blank(),
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm")
  ) +
  guides(fill = guide_legend(nrow = 2, override.aes = list(shape = 22, size = 4))) +
  labs(title = "Mutation signature analysis across three cohorts") +
  stat_pvalue_manual(
    data = wilcox.res,
    tip.length = 0.01,
    xmin = "group1",
    xmax = "group2",
    label = "p.adj.signif",
    label.size = 2,
    inherit.aes = FALSE,
    hide.ns = TRUE
)
```

### Session Information

```{r}
sessionInfo()
```